---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cyclinbayes

<!-- badges: start -->
<!-- badges: end -->

Cyclinbayes is an R package implementing Bayesian methods for estimating both directed acyclic graphs (DAGs) and directed cyclic graphs (DCGs). The package provides full posterior inference for graph structures and causal effects using a hierarchical Bayesian model, allowing principled uncertainty quantification, edge inclusion probabilities, and credible intervals. 

For DAGs, cyclinbayes uses a hybrid MCMC scheme that combines collapsed Gibbs sampling with simulated annealing to improve mixing and avoid local optima. For DCGs, the package performs joint updates of adjacency and causal effect coefficients using random walk proposals, enabling inference in systems with feedback cycles.For DCGs, the package performs joint updates of adjacency and causal-effect coefficients using random-walk proposals, enabling inference in systems with feedback cycles. In both settings, sparsity is effectively recovered using spike and slab priors. 

Implemented in Rcpp, cyclinbayes leverages optimized C++ routines to handle large-scale, high-dimensional datasets. 

## Installation from Github
You can install the development version of cyclinbayes from GitHub using remotes:

```{r}
#install.packages("remotes")
#remotes::install_github("roblee01/cyclinbayes")
```

Then load the package:

```{r}
library(cyclinbayes)
library(ggplot2)
library(SID)
library(pcalg)
```

## Example

Below is a simple example demonstrating how to use the Bayesian LiNGAM (DAG) sampler. Let \(p\) denote the number of variables and \(n\) the sample size. We generate error terms from a finite Gaussian mixture model:

\[
\epsilon_i^{(q)} \sim \sum_{k=1}^{M} \pi_{ik} N(\mu_{ik}, \tau_{ik}),
\]

with mixture components

\[
M = 2, \quad 
(\mu_{i1}, \mu_{i2}) = (-0.5, 0.5), \quad
(\tau_{i1}, \tau_{i2}) = (0.1, 0.3), \quad
(\pi_{i1}, \pi_{i2}) = (0.5, 0.5).
\]

We generate a sparse causal-effect matrix \(B\) by sampling edges independently with
probability \(\Delta = 0.9\) until the adjacency matrix is acyclic.

Given \(B\) and the error matrix \(\epsilon\), the data are generated via

\[
Y = (I - B)^{-1}\epsilon,
\]

where the \(i\)-th row of \(Y\) corresponds to

\[
(Y_i^{(1)}, \ldots, Y_i^{(n)})^\top.
\]

```{r example}
#set.seed for reproducibility
set.seed(21)


#######################################
# Simulation and MCMC settings
#######################################
N              <- 300    # sample size
num_covariates <- 10     # number of features
M              <- 2      # number of mixture components
num_iter       <- 10000  # number of MCMC iterations

#######################################
# Hyperparameter setup
#######################################
params = list(
  a_mu      = 0,
  b_mu      = 2,
  a_gamma   = 0.5,
  b_gamma   = 0.5,
  a_gamma_1 = 2,
  b_gamma_1 = 1,
  a_tao     = 2,
  b_tao     = 1,
  a_og_tao  = 0.01,
  b_og_tao  = 0.01,
  alpha     = 1
)

#######################################
# Generate DAG example data
#######################################
example_list = generates_examples_DAG(
  num_covariates = num_covariates,
  N = N,
  M = M,
  prob_sparsity = 0.9,
  seed_input = 21
)

data_matrix = example_list$data_matrix
Adjacency_matrix_true = example_list$Adjacency_matrix_true

#######################################
# Run Bayesian LiNGAM (DAG) sampler
#######################################
results_lists = BayesDAG(
  data_matrix,
  params$a_mu,
  params$b_mu,
  params$a_gamma,
  params$b_gamma,
  params$a_tao,
  params$b_tao,
  params$a_og_tao,
  params$b_og_tao,
  params$a_gamma_1,
  params$b_gamma_1,
  params$alpha,
  M,
  num_iter
)

#######################################
# Extract key posterior summaries
#######################################
Adjacency_matrix_means = results_lists$Adjacency_matrix_means 
Adjacency_matrix_list = results_lists$Adjacency_matrix_list 
Causal_effect_matrix_list = results_lists$Causal_effect_matrix_list
gamma_list = results_lists$gamma_list
gamma_1_list = results_lists$gamma_1_list
mu_matrix_list = results_lists$mu_matrix_list
tao_matrix_list = results_lists$tao_matrix_list
pi_matrix_list = results_lists$pi_matrix_list
```

To obtain a representative estimate of the graph structure, we use the function `posterior_adjacency_analysis()`, which selects the posterior weighted medoid under a chosen distance metric. For DCGs, we consider the Structural Hamming
Distance (SHD) and the Frobenius norm; the Structural Intervention Distance (SID) is only applicable when the sampled graphs are DAGs.

```{r}
Adjacency_matrix = select_posterior_graph(Adjacency_matrix_list,dist_type = 'shd')
Adjacency_matrix
```

```{r, warning=FALSE, message=FALSE}
Adjacency_matrix = select_posterior_graph(Adjacency_matrix_list,dist_type = 'sid')
Adjacency_matrix
```

```{r}
Adjacency_matrix = select_posterior_graph(Adjacency_matrix_list,dist_type = 'forb')
Adjacency_matrix
```
All three selected graphs match the true adjacency matrix:

```{r}
Adjacency_matrix_true
```
We can also examine posterior uncertainty for individual parameters. For instance, below we plot the posterior samples of the edge inclusion probability $\gamma$:


```{r}
plot(gamma_list[,1][(0.75*num_iter):num_iter],type='l', ylab='gamma value')
```
To summarize this parameter, we use summary_posterior_vec() to obtain the 95% credible interval and highest posterior density (HPD):

```{r}
gamma_posterior_summary = summary_posterior_vec(gamma_list,0.95)
gamma_posterior_summary$credible_interval
gamma_posterior_summary$hpd_interval[,1]
```
Next, we illustrate posterior uncertainty for the causal effect coefficients. Below we show the HPD and credible intervals for the nonzero entries of the causal effect matrix, with the dashed red line indicating the true weight. The credible intervals tightly capture the true values, and the HPD estimates lie close to the reference line, demonstrating that the method recovers both causal strength and graph structure accurately.

```{r}
Causal_effect_matrix_summary = summary_posterior_matrix(Causal_effect_matrix_list, level = 0.95)
hpd_matrix_acyclic = Causal_effect_matrix_summary$hpd_matrix
ci_matrix_acyclic = Causal_effect_matrix_summary$ci_matrix

#######################################
# Extracting nonzero HPD intervals
#######################################
par(mfrow=c(2,1))

nonzero_cols = which(colSums(hpd_matrix_acyclic) != 0)
num_non_zero_coef = length(nonzero_cols)
data_1 = data.frame(cbind(1:num_non_zero_coef,t(hpd_matrix_acyclic[,which(colSums(hpd_matrix_acyclic)!=0)])))

ggplot(data_1) + 
  geom_segment(aes(x = lower, xend = upper, y = V1, yend = V1)) + 
  geom_vline(xintercept = 1, linetype = "dashed") + 
  labs(y = NULL, x = "HPD interval") + 
  theme_minimal()


#######################################
# Credible intervals
#######################################
data_2 = data.frame(ci_matrix_acyclic[which(rowSums(ci_matrix_acyclic)!=0),])
x = 1:nrow(data_2)
data_2 = cbind(x,data_2)

ggplot(data_2, aes(x = x, y = X2)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  geom_errorbar(aes(ymin = X1, ymax = X3), width = 0.2) +  # just X1/X3
  labs(y = "Weight Estimate with 95% CI", x = "X") +
  theme_minimal()

```

Finally, to illustrate the cyclic Bayesian sampler, we first generate an adjacency matrix that contains at least one cycle. To ensure that the matrix \(I - B\) is invertible and the system remains stable, we constrain the spectral radius of the causal-effect matrix \(B\).

Specifically, we compute
\[
\rho(B) = \max \{ |\lambda| : \lambda \in \mathrm{eig}(B) \},
\]
and if \(\rho(B) \ge 0.95\), we rescale the matrix as
\[
B \leftarrow \frac{0.95}{\rho(B)} \, B,
\]
which guarantees that the spectral radius is strictly below \(0.95\).

This rescaling step ensures numerical stability of the likelihood while preserving the relative structure of the causal effects encoded in \(B\).

```{r}
library(cyclinbayes)

#######################################
# Simulation and MCMC settings
#######################################
N = 250 # Sample size for the test data
num_covariates = 7 # Number of features for test data
M = 2 # Number of finite clusters for mixed normal in likelihood
num_iter = 10000 # Number of iterations MCMC runs

#######################################
# Hyperparameter setup
#######################################
params = list(
  a_mu      = 0,
  b_mu      = 2,
  a_gamma   = 2,
  b_gamma   = 1,
  a_gamma_1 = 2,
  b_gamma_1 = 1,
  a_tao     = 2,
  b_tao     = 1,
  alpha     = 1
)

#######################################
# Generate DCG example data
#######################################

example_list = generates_examples_DCG(
  num_covariates,
  N,
  M,
  0.9,
  21
)

data_matrix = example_list$data_matrix
Adjacency_matrix_true = example_list$Adjacency_matrix_true
Causal_effect_matrix_true = example_list$Causal_effect_matrix_true

###############################################################################

results_list = BayesDCG(
  data_matrix,
  params$a_mu,
  params$b_mu,
  params$a_gamma,
  params$b_gamma,
  params$a_tao,
  params$b_tao,
  params$a_gamma_1,
  params$b_gamma_1,
  params$alpha,
  M,
  num_iter
)

#######################################
# Extract posterior outputs
#######################################
Adjacency_matrix_means = results_list$Adjacency_matrix_means 
Adjacency_matrix_list = results_list$Adjacency_matrix_list 
Causal_effect_matrix_list = results_list$Causal_effect_matrix_list
gamma_list = results_list$gamma_list
gamma_1_list = results_list$gamma_1_list
mu_matrix_list = results_list$mu_matrix_list
tao_matrix_list = results_list$tao_matrix_list
pi_matrix_list = results_list$pi_matrix_list
```

To obtain a representative estimate of the graph structure, we use the function `posterior_adjacency_analysis()`, which selects the posterior weighted medoid under a chosen distance metric. For DCGs, we consider the Structural Hamming Distance (SHD) and the Frobenius norm; the Structural Intervention Distance (SID) is only applicable when the sampled graphs are DAGs.

```{r,warning=FALSE, message=FALSE}
# SID is included here for completeness; it is not applicable to general DCGs
select_posterior_graph(Adjacency_matrix_list, dist_type = 'sid')
```
```{r}
select_posterior_graph(Adjacency_matrix_list,dist_type = 'shd')
```
```{r}
select_posterior_graph(Adjacency_matrix_list,dist_type = 'forb')
```
In this example, the selected graphs coincide with the true adjacency matrix:
```{r}
Adjacency_matrix_true
```
We can also examine individual parameters, such as the edge inclusion probability $\gamma$. Below we plot the MCMC samples for $\gamma$ (after the first 75% of iterations) to assess mixing and posterior behavior:

```{r}
plot(gamma_list[,1][(0.75*num_iter):num_iter],type='l',ylab='gamma value')
```

We can summarize this parameter using `summary_posterior_vec()`, which returns both the 95% credible interval and the highest posterior density (HPD) interval:

```{r}
# 95% credible interval + HPD interval
gamma_posterior_summary = summary_posterior_vec(gamma_list, 0.95)

# Print equal tailed credible interval
gamma_posterior_summary$credible_interval

# Print HPD interval (first column corresponds to gamma)
gamma_posterior_summary$hpd_interval[,1]
```

Similarly, we can summarize uncertainty in the causal effect coefficients using summary_posterior_matrix(), which computes HPD and equal tailed credible intervals for each entry of the posterior causal effect matrix:

```{r}
# Compute HPD and CI summaries for causal effect matrix
Causal_effect_matrix_summary = summary_posterior_matrix(Causal_effect_matrix_list, level = 0.95)
hpd_matrix_cyclic = Causal_effect_matrix_summary$hpd_matrix
ci_matrix_cyclic = Causal_effect_matrix_summary$ci_matrix

#######################################
# Extracting nonzero HPD intervals
#######################################

par(mfrow=c(2,1))
num_non_zero_coef = ncol(hpd_matrix_cyclic[,which(colSums(hpd_matrix_cyclic)!=0)])
data_1 = data.frame(cbind(1:num_non_zero_coef,t(hpd_matrix_cyclic[,which(colSums(hpd_matrix_cyclic)!=0)])))


ggplot(data_1) +
  geom_segment(aes(x = lower, xend = upper, y = V1, yend = V1)) +
  geom_vline(xintercept = 0.7143305, linetype = "dashed") +
  labs(y = NULL, x = "HPD interval") +
  theme_minimal()

#######################################
# Credible intervals
#######################################

data_2 = data.frame(ci_matrix_cyclic[which(rowSums(ci_matrix_cyclic)!=0),])
x = 1:nrow(data_2)
data_2 = cbind(x,data_2)

ggplot(data_2, aes(x = x, y = X2)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.7143305, linetype = "dashed", color = "red") +
  geom_errorbar(aes(ymin = X1, ymax = X3), width = 0.2) +  
  labs(y = "Weight Estimate with 95% CI", x = "X") +
  theme_minimal()

```




























